Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-04 tagless/tagless-final.tex
start crafting some language on related work

\section{Related work}\label{related}
Walid's Jones optimality paper. See his own ICFP02 paper that
described what he achieved then and what he \emph{did not}. He did not
eliminate all the tags. Ken's and mine impression of it:
(i) we can't write a \emph{direct} interpreter of the type language
           -- the only known to them, and \emph{partial} solution, was 
              that by Zhe Yang;
(ii) we have to use indirect interpreter, which needs the universal
           type and hence tagging;
(iii) that means any self-interpreter must have tags
(iv) that means not Jones-optimal.
It seems we should attack the point1 (in which case we don't have to
bother with self-interpreters). We say that we \emph{can} write the direct
interpreter, and without GADT or their equivalents (typeclasses).
Meta-language computes types, and the compiler and interpreter respect
those types and needs no tags.

Walid's ICFP02

Walid's PEPM07 (Concoqtion paper)? At leats at the Concoqtion
presentation at WG2.11, Emir used tagless interpreter as the main and
only example. \jacques{but at the PEPM07 presentation, Walid did not
quite do it that way.  He presented Concoqtion more as an exploration
of (an obvious but difficult) part of the design space for programming
languages.}

Danvy's self-interpreter
Ken has read Danvy et al's paper and said that the paper
doesn't seem to even make the claim that the source language is
typed. The interpreter is typed, but the source language doesn't seem
to be.
In sharp contrast to Danvy's paper, all the evaluators in our case
cannot produce any pattern-match error! Except for the partial
evaluator, there are NO patterns to match. And this is critical for
tag elimination -- pattern matching means there are no tags. And in
PE, the pattern-match deals with phases rather than with types. Also,
in our case the exhaustiveness of the pattern match is apparent to the
compiler -- that's why there are no tags.

Simon Peyton-Jones GADTs (the motivation for GADT is tagless
interpreter, but only for the first-order language)

Xi's POPL03 (He uses a higher-order language, one of the motivations
for his GRDT)

>>>>

\section{Related work}\label{related}

Our initial motivation came from several papers 
\citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} which used
embedded interpreters as a central example to justify looking 
into advanced type systems.  While we admired all this technical work,
there remained the issue that we were not convinced of the strength
of their motivating example.

\citet{WalidICFP02}, when referring to the work of \citet{taha-tag},
claims that this shows that self-interpretation is possible in a
\emph{simply typed $\lambda$-calculus}, but this is somewhat misleading
as their type system does not differentiate between the various ground
types.  In other words, various constants (like \textsf{nil} and 
\textsf{true}) both inhabit the same type $D$.  This is not very satisfying.

\oleg{He did not eliminate all the tags.}\jacques{can you show me?}
From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
impression is that the claim is:
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type,
\item It is not possible to write a \emph{direct} interpreter for an typed
language encoded this way, in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus we have to use an indirect interpreter, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the first step
was not valid.

\oleg{the only known to them, and \emph{partial} solution, was 
  that by Zhe Yang;}\jacques{I cannot find that reference}

\citet{Danvy-tagging-encoding} discusses Jones optimality at length.
There, the authors demonstrate the usefulness of HOAS for typed
self-interpretation.  However, no claim is made that the source program
is itself typed -- just that the interpreter is.  Our reading is that
the source language is not typed.  However, in sharp contrast with 
their interpreter, none of our evaluators can product pattern-match errors
(unlike the code that accompanies \citet{Danvy-tagging-encoding}).
In other words, their language (and interpreter) does use tags, however
the shift to HOAS allows the partial evaluator to remove all the tags.
In contrast, we have no tags at all.
